A Process for the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE)

### 2025-02-16 manuscript/manuscript.qmd
added Ontada comments

[**Competing Interests Statement:**]{.underline} Dr. Weberpals is now an employee of AstraZeneca and owns stocks in AstraZeneca. Dr. Wang has consulted ad hoc for Exponent Inc. and MITRE a federally funded research center for the Centers for Medicare and Medicaid Services on unrelated work. Dr. Schneeweiss is participating in investigator-initiated grants to the Brigham and Women's Hospital from Boehringer Ingelheim, Takeda, and UCB unrelated to the topic of this study. He owns equity in Aetion Inc., a software manufacturer. He is an advisor to Temedica GmbH, a patient-oriented data generation company. His interests were declared, reviewed, and approved by the Brigham and Women's Hospital in accordance with their institutional compliance policies.

@@

Real-world evidence (RWE) studies are increasingly used to complement evidence from randomized controlled trials (RCTs), contextualizing the effectiveness and safety of medical interventions as delivered in clinical practice. Advancements in the curation and accessibility of electronic health record data (EHR) have presented the opportunity to investigate disease domains such as oncology, where administrative healthcare claims databases alone are not fit-for-purpose. The RCT DUPLICATE initiative has previously enhanced understanding of when RWE studies can draw causal conclusions by emulating trials in the cardio-metabolic, renal, and pulmonary clinical areas. In this process paper, we present the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which aims to extend this work to oncology. This expansion will emulate 12 RCTs in multiple EHR databases across four different cancer indications, including non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma. The objectives of this project are to develop state-of-the-art methodological approaches and to apply them in a demonstration project designed to create insights for guiding the use of RWE in oncology. The project will place a special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to analysis. Agreement of treatment effect estimates between RCTs and their respective emulations will be assessed using pre-specified criteria. Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project will inform understanding of how measurement, design, and analytic decisions influence bias and validity in oncological RWE studies.

@@

Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products. With the advent of the 21^st^ Century Cures Act directive,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023[@senior2024fresh], RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology where potential use cases include the assessment of effectiveness in patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.

However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors and frequently referenced limitations include missing data, small sample sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in guidelines in real-world care and the inability to measure and emulate common eligibility criteria, prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of different oncology trials in multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies and to provide context as to which questions can be validly answered.

@@

The focus of ENCORE is to maximize potential learnings on when RWE studies can or cannot yield similar results compared to RCTs. To that end, the emphasis of the project is on trials of therapies for the most common cancers for which there has been substantial therapeutic development in recent years. After careful review and exchange with clinical and regulatory experts, four cancer indications were identified including lung cancer, breast cancer, colorectal cancer and multiple myeloma. For each cancer type we aim to conduct three trial emulations using multiple databases accessible for the scope of this project (i.e., the total number of emulations will equal 12 trials x *n* databases which are found fit-for-purpose for each trial).

We used a semi-automated process for trial selection where the eligibility criteria are documented in CONSORT diagrams showing reasons for excluding RCTs for each cancer type. The search was conducted using the AACT database which is a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) that contains all information (protocol and result data elements) about every study registered on ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we used a combined search query strategy of the National Library of Medicine (NLM)-controlled *MeSH* term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.

@@

-   Feasibility and clinical relevance

@@

A list of tentative, shortlisted primary candidates is presented in @tbl-rcts and the corresponding selection process is illustrated in the CONSORT diagrams (Supplementary Figures 1-4). Naturally, the majority of trials cover advanced or metastatic cancer populations because a large proportion of drug development efforts have focused on these settings in recent years. A key learning that we aim to foster with the shortlisted trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and population characteristics (simple versus complex genetic or immunological signatures) can be emulated using RWD. If more thorough feasibility assessments suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the sample size is not sufficient, then runner-up candidates will be considered instead.

@@

The ENCORE project will use data from four oncology-specific electronic health records (EHR)-derived data sources (in alphabetical order): ConcertAI, COTA, Flatiron Health, McKesson/Ontada. All available databases draw from a comprehensive national sample of patients with cancer in the US with detailed EHR-derived information on the information necessary to study medication effectiveness in oncology. A detailed description and methodology on how patients are sampled will be provided with each trial emulation protocol. For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (the numbering does not coincide with the above order of mention of the databases). If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects (weights reflect the inverse of the variance of each database study) and random effects (weights reflect the inverse of the variance of each database study plus an additional variance term that quantifies the assumed heterogeneity between databases) models.[@nikolakopoulou2014interpret]

@@

For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] which has been recommended for regulatory submissions of RWE studies,[@guidelinegeneral] and will be registered on ClinicalTrials.gov after careful review by a clinical and FDA regulatory expert panel. Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to (see @sec-agreement-metrics). All eligibility criteria will be extracted based on publications, publicly available protocols and statistical analysis plans of the selected RCT.

@@

Real-world data fitness and emulation feasibility for each shortlisted candidate trial will be assessed in multiple steps based on guidance of the oncology quality, characterization, and assessment of real-world data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure/line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care. Since a vast majority of oncological RCTs in recent years have focused on selected, biomarker-defined populations, subtleties in measurement and operationalizability of specific biomarkers must be reflected to ensure a representative and large enough study population. For example, immunotherapies have significantly changed the cancer treatment landscape since the approval of the first PD-L1 inhibitor in 2015. With many trials that have followed thereafter, the operationalization of the expression of the PD-L1 biomarker in RCTs (e.g., as a percent staining, tumor proportion score or combined positive score) has also evolved since then and PD-L1 '*positivity'* may have different definitions across calendar years based on different cut-off values.

According to the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], the next step will outline tables that describe how eligibility criteria will be ascertained using a color-coded heatmap that will indicate the level of confidence on how well each criterion can be emulated in each selected database. As there are general eligibility criteria in oncology trials which either will not be possible to emulate (e.g., physician-assessed survival prognosis of xy months) or that are clinically not relevant for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will decide on key eligibility criteria for the emulation of the trial.

We will additionally provide a definition on how exposures, outcomes and covariates are exactly defined and operationalized in each respective database. There will be special emphasis on how exposure, in context of their respective disease and line of therapy settings, and the OS outcome will be emulated. For all considered databases, the OS endpoint is typically a composite that is derived from different sources comprising EHR abstractions, social security death index, obituary and other linkages. Given that not all relevant sources that provide mortality data are synchronized and updated uniformly, sensitivity analyses with more conservative (i.e., earlier) censoring dates will be considered for each trial emulation to mitigate the potential impact of ghost-time bias.[@meyer2020open]

@@

Causal analyses of observational data may not have the same pre-requisites in terms of formal hypothesis testing and statistical power than RCTs since the number of 'recruited' patients is given and cannot be influenced.[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreements metrics between RCT and RWD results. Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients. To assess if the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power (1-$\beta$) will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]

@@

To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR and corresponding 95% confidence intervals), we will derive three qualitative agreement metrics: statistical agreement, estimate agreement and agreement based on the standardized mean difference (SMD). Examples are illustrated @tbl-metrics.

@@

**Estimate agreement**: agreement that the estimated RWE treatment effect is within the 95% CI of the RCT treatment effect estimate. Provided that for some emulations, the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there is no regulatory agreement although the treatment effect estimates are highly overlapping but with the RCT estimate crossing the null (or vice versa in case the RCT has a larger power than the RWE emulation).

@@

where $\hat{\theta}_{RCT}$ and $\hat{\theta}_{RWE}$ are the treatment effect estimates (hazard ratios or median survival times) and $\text{Var}(\hat{\theta}_{RCT})$ and $\text{Var}(\hat{\theta}_{RWE})$ are the corresponding variances for RCT and RWE, respectively. The resulting SMDs will be interpreted such that with an SMD of 1.00, the effect estimate from the RCT and the RWE emulation are 1 standard deviation apart. For an $\alpha$-level of 0.05, the null hypothesis of no difference would be rejected whenever $|Z| > 1.96$.

@@

To establish an analytic cohort, key eligibility criteria will be applied in which patients with missing values in eligibility criteria are considered eligible in the respective attrition steps to allow for thorough missing data investigations. These missing data investigations will empirically assess assumptions on potentially underlying missingness mechanisms according to Rubin’s classification of missing data (i.e., missing completely at random \[MCAR\], missing at random \[MAR\] and missing not at random \[MNAR\]).[@rubin1976inference] To that end, we will adopt a principled process on missing data that was developed as part of a FDA Sentinel Innovation Center causal inference workstream that empirically evaluates different aspects across partially observed covariates based on three group diagnostics.[@weberpals2024smdi; @weberpals2024] In brief, these diagnostic cover (1) comparisons of patients characteristics with and without an observed level of the partially observed covariate, (2) ability to predict missingness given observed data, and (3) assessments if outcomes between patients with a missing value are systematically different. Together with expert domain knowledge and assumptions about the underlying missing data structure through canonical causal diagrams,[@moreno2018canonical] this will inform decisions regarding the in- or exclusion of patients with missing values in key eligibility criteria and potential sensitivity analyses to assess the robustness of these decisions.[@tompsett2018use]

@@

Building on an established approach based on the RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized], the ENCORE project aims to emulate 12 oncology trials using multiple EHR data sources to inform the potential use of RWE for regulatory science in oncology. The project focuses on four common cancers and will evaluate the agreement of treatment effect estimates between RCTs and their respective emulations. Historically, administrative health claims databases have been the backbone of most research in RWE. With increasing access to EHR and a maturing set of methodological approaches to draw causal inferences from such data, the ENCORE project is one of the first of its kind to evaluate when and how RWD can be used to deliver similar causal conclusions compared to RCTs in the field of oncology.

The RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized] identified multiple emulation challenges which we expect to also encounter in ENCORE. One particular aspect that we may not always be able emulate is the exact distribution of patient characteristics of the trial population. This can be due to the lack of data granularity and comprehensiveness to emulate relevant eligibility criteria or the fact that large pivotal trials in oncology are typically conducted in multiple countries worldwide, which we will not be able to mirror given that all considered databases reflect the US only. This may be a critical factor especially given that the pathophysiology, prognosis and factors that drive heterogeneous treatment outcomes of certain cancers differ between countries. For example, some cancer indications (e.g., GI cancers) or some genetic mutations (e.g., EGFR mutations) are much more prevalent in Asian countries compared to the US or Europe.

@@

This issue may be even augmented with more complex treatment protocols like combination regimens (e.g., CheckMate9LA) as compared to monotherapies (e.g., CheckMate057) since the ascertainment of the exposure needs to happen over a pre-defined time window which is often difficult to calibrate in RWD. As a result, exposure misclassification (due to too narrow ascertainment windows) and selection bias (due to too long ascertainment windows) are common trade-offs in such scenarios. Alternative analytic approaches which target a per-protocol estimand, such as the clone-censor-weight design,[@gaber2024mystifying] may be viable options if the ITT estimand cannot be estimated due to the aforementioned parameters, provided that necessary covariate measurements are available to account for the artificial censoring introduced with these methods.

@@

```{r}
#| label: fig-initiators
#| fig-cap: "Visualization of descriptive drug utilization analyses displaying a) initiation trends between compared regimens based on calendar time, b) cumulative rate of patients switching to another line of treatment."
#| fig-widh: 3
#| fig-height: 6

>>>>

[**Competing Interests Statement:**]{.underline} Dr. Weberpals is now an employee of AstraZeneca and owns stocks in AstraZeneca. Dr. Kehl has received research funding from Meta, Inc. to his institution. Drs. Espirito and Robert are employees of McKesson and own McKesson stock. Dr. Wang has consulted ad hoc for Exponent Inc. and MITRE a federally funded research center for the Centers for Medicare and Medicaid Services on unrelated work. Dr. Schneeweiss is participating in investigator-initiated grants to the Brigham and Women's Hospital from Boehringer Ingelheim, Takeda, and UCB unrelated to the topic of this study. He owns equity in Aetion Inc., a software manufacturer. He is an advisor to Temedica GmbH, a patient-oriented data generation company. His interests were declared, reviewed, and approved by the Brigham and Women's Hospital in accordance with their institutional compliance policies.

@@

Real-world evidence (RWE) studies are increasingly used to complement evidence from randomized controlled trials (RCTs), contextualizing the effectiveness and safety of medical interventions as delivered in clinical practice. Advancements in the curation and accessibility of electronic health record data (EHR) have presented the opportunity to investigate disease domains such as oncology, where administrative healthcare claims databases alone are not fit-for-purpose. The RCT DUPLICATE initiative has previously enhanced understanding of when RWE studies can draw causal conclusions by emulating trials in the cardio-metabolic, renal, and pulmonary clinical areas. In this process paper, we present the Emulation of Comparative Oncology Trials with Real-world Evidence (ENCORE) project, which aims to extend this work to oncology. This expansion will emulate 12 RCTs in multiple EHR databases across four different cancer indications, specifically non-small cell lung cancer, breast cancer, colorectal cancer, and multiple myeloma. The objectives of this project are to develop state-of-the-art methodological approaches and to apply them in a demonstration project designed to create insights for guiding the use of RWE in oncology. The project will place a special emphasis on systematic evaluation of fitness of data in relation to the study design and statistical analysis for a particular research question, and pre-registration of study protocols prior to analysis. Agreement of treatment effect estimates between RCTs and their respective emulations will be assessed using pre-specified criteria. Through extensive sensitivity analyses benchmarked against RCT results, the ENCORE project will inform understanding of how measurement, design, and analytic decisions influence bias and validity in oncological RWE studies.

@@

Randomized controlled trials (RCTs) have been the gold standard for establishing the efficacy and safety of medical products. With the advent of the 21^st^ Century Cures Act directive,[@RWEFDA] the Food and Drug Administration (FDA) established a framework to increasingly consider real-world evidence (RWE) generated from routine-care health data such as electronic health records (EHR) to evaluate and contextualize the comparative safety and effectiveness of novel cancer therapies.[@purpura2022role] Accounting for 21% of all approvals, oncology was the disease area with the most FDA drug approvals in 2023.[@senior2024fresh] RWE has particularly important potential to complement evidence coming from RCTs in the field of precision oncology, where potential use cases include the assessment of effectiveness in rare patient populations that are underrepresented in RCTs, the construction of external control arms in single-arm trials where active recruitment to a randomized trial may not be feasible, or the discovery of biomarkers among pan-tumor populations that harbor specific genomic and immuno-pathological signatures.

However, the validity and transportability of results derived between RWE studies and RCTs can depend on many factors. Frequently referenced limitations include missing data, small sample sizes, data discontinuity,[@Merola2022; @joshua2022longitudinal] adoption of changes in guidelines in real-world care and the inability to measure and emulate common eligibility criteria, including prognostic factors, and standardized response assessments in real-world data (RWD).[@rider2024emulations] While there are already published examples of oncology trial emulations,[@rider2024emulations; @merola2023aetion; @merola2024calibrating] a systematic and scaled approach to emulate a diverse set of different oncology trials with multiple heterogeneous databases is necessary to gain confidence in the validity of RWE studies and to provide context as to which questions can be validly answered.

@@

The focus of ENCORE is to maximize potential learnings on when RWE studies can or cannot yield similar results compared to RCTs. To that end, the emphasis of the project is on trials of therapies for the most common cancers for which there has been substantial therapeutic development in recent years. After careful review and exchange with clinical and regulatory experts, four cancer indications were identified: non-small cell lung cancer, breast cancer, colorectal cancer and multiple myeloma. For each cancer type, we aim to conduct three trial emulations using multiple databases accessible for the scope of this project (i.e., the total number of emulations will equal 12 trials x *n* databases which are found fit-for-purpose for each trial).

We used a semi-automated process for trial selection where the eligibility criteria are documented in CONSORT diagrams showing reasons for excluding RCTs for each cancer type. The search was conducted using the *Aggregated Analysis of ClinicalTrials.gov* (AACT) database which is a publicly available relational database developed and maintained by the Clinical Trials Transformation Initiative (CTTI) that contains all information (protocol and result data elements) about every study registered on ClinicalTrials.gov.[@tasneem2012database] To identify eligible trials, we used a combined search query strategy of the National Library of Medicine (NLM)-controlled *Medical Subject Headings **(***MeSH) term and a free keyword search for the respective cancer indication in the *conditions*, *studies* and *detailed_descriptions* fields of each trial entry on ClinicalTrials.gov.

@@

-   Feasibility and clinical relevance (latter was defined as treatment or paradigm-changing trials or such that challenged existing treatment policies)

@@

A list of tentative, shortlisted primary candidates is presented in @tbl-rcts and the corresponding selection process is illustrated in the CONSORT diagrams (Supplementary Figures 1-4). Naturally, the majority of trials cover advanced (locally advanced, inoperable, recurrent/progressive disease) or metastatic cancer populations because a large proportion of drug development efforts start in these settings. A key objective that we aim to foster with the shortlisted trials is to achieve a better understanding how different disease settings (early, late), line settings (\[neo\]adjuvant, first line, advanced lines of therapy), therapy protocols (monotherapy, combination therapy) and population characteristics (simple versus complex genetic or immunological signatures) can be emulated using RWD. If more thorough feasibility assessments suggest that the threat of bias from mis-measurement of key study parameters or residual confounding remains high, or that the sample size is not sufficient, then runner-up candidates will be considered instead.

@@

The ENCORE project will use data from four oncology-specific electronic health records (EHR)-derived data sources (in alphabetical order): ConcertAI, COTA, Flatiron Health, and Ontada/McKesson. All available databases draw from a comprehensive national sample of patients with cancer in the US with the detailed EHR-based data necessary to study medication effectiveness in oncology. A detailed description and methodology on how patients are sampled will be provided with each trial emulation protocol. For ENCORE, not all databases will be available for each cancer indication and the names of the databases will be blinded and referred to as ENCORE DataBase (EDB) 1, 2, 3 and 4 for the final reporting of results (the numbering does not coincide with the above order of mention of the databases). If more than one database is considered fit-for-purpose for a respective trial emulation, the best possible analytic model will be employed for each database separately and final treatment effect estimates will be pooled using a meta-analytic approach using fixed effects (weights reflect the inverse of the variance of each database study) and random effects (weights reflect the inverse of the variance of each database study plus an additional variance term that quantifies the assumed heterogeneity between databases) models.[@nikolakopoulou2014interpret]

@@

For each shortlisted and selected RCT, a detailed protocol, pre-specifying key elements of the trial emulation, will be developed using the HARPER protocol template[@wang2022harmonized] recommended for regulatory submissions of RWE studies[@guidelinegeneral] and will be registered on ClinicalTrials.gov after careful review by a clinical and FDA regulatory expert panel. Following the target trial emulation framework, we will provide an explicit statement and rationale on how each element will be emulated including database selection, covariate measurement, operationalization of key eligibility criteria, study design, data analysis and causal contrasts of interest.[@hernan2022target; @hernan2016specifying] Since it is common that oncology RCTs update OS estimates periodically based on accrued follow-up time, the protocol will give a brief summary of each emulated RCT and specify which target OS estimates will be used to compare agreement metrics to (see @sec-agreement-metrics). All eligibility criteria will be extracted based on publications, publicly available protocols and statistical analysis plans of the selected RCT.

@@

Real-world data fitness and emulation feasibility for each shortlisted candidate trial will be assessed in multiple steps based on guidance of the oncology quality, characterization, and assessment of real-world data (Oncology QCARD) Initiative.[@rivera2022friends] The first step assesses if relevant variables like exposure/line of therapy, outcomes, and covariates are generally available, measured and operationalizable in routine-care. Since many oncological RCTs in recent years have focused on selected, biomarker-defined populations, subtleties in measurement and operationalizability of specific biomarkers must be reflected to ensure a representative and large enough study population. For example, immune checkpoint inhibitors have significantly changed the cancer treatment landscape since the first approvals of the CTLA-4 inhibitor ipilimumab in 2011 and the PD-1 inhibitors pembrolizumab and nivolumab in 2014 in the United States. With many trials that have followed thereafter, the operationalization of the expression of the PD-L1 biomarker in RCTs (e.g., as a percent staining, tumor proportion score or combined positive score) has also evolved, and PD-L1 '*positivity'* may have different definitions across calendar years based on different cut-off values.

According to the *Structured Process to Identify Fit-For-Purpose Data* (SPIFD) framework[@gatto2022structured], the next step will outline tables that describe how eligibility criteria will be ascertained using a color-coded heatmap that will indicate the level of confidence on how well each criterion can be emulated in each selected database. As there are general eligibility criteria in oncology trials which either will not be possible to emulate (e.g., physician-assessed survival prognosis of x months) or that are clinically not relevant for the emulation of the trial (e.g., male patients should be willing to use barrier contraception), the study team will decide on key eligibility criteria for the emulation of the trial.

We will additionally provide a definition on how exposures, outcomes and covariates are exactly defined and operationalized in each respective database. There will be special emphasis on how exposure, in context of their respective disease and line of therapy settings, and the OS outcome will be emulated. For all considered databases, the OS endpoint is typically a composite that is derived from different sources comprising EHR documentation, social security death index, obituary and other linkages. Given that not all relevant sources that provide mortality data are synchronized and updated uniformly, sensitivity analyses with more conservative (i.e., earlier) censoring dates will be considered for each trial emulation to mitigate the potential impact of ghost-time bias.[@meyer2020open]

@@

Causal analyses of observational data may not have the same pre-requisites in terms of formal hypothesis testing and statistical power as RCTs since the number of 'recruited' patients is given and cannot be influenced.[@hernan2022causal] For this project, however, the emulations in EHR data must have at least equal power to the relevant trial for interpretation of the pre-specified agreements metrics between RCT and RWD results. Since the main outcome of interest is defined as time to all-cause mortality (OS), the estimation of the statistical power is driven by the number of events rather than the number of patients. To assess if the overall number of events, unstratified by exposure, is sufficient such that a significant difference can be detected based on the original RCT-reported hazard ratio (HR), the statistical power (1-$\beta$) will be estimated using Schoenfeld's sample-size formula for the proportional-hazards regression model.[@schoenfeld1981asymptotic]

@@

To formally compare treatment effects between RCTs and their respective emulations, we will adapt the approach of the RCT-DUPLICATE project.[@franklin2020nonrandomized; @wang2023emulation] That is, for the primary endpoint of interest (HR for OS and corresponding 95% confidence intervals), we will derive three qualitative agreement metrics: statistical agreement, estimate agreement and agreement based on the standardized mean difference (SMD). Examples are illustrated @tbl-metrics.

@@

**Estimate agreement**: agreement that the estimated RWE treatment effect is within the 95% CI of the RCT treatment effect estimate. Provided that for some emulations, the power of the RWE study may be larger than that of the original RCT, this could lead to situations where there is no statistical significance agreement although the treatment effect estimates are highly overlapping but with the RCT estimate crossing the null (or vice versa in case the RCT has a larger power than the RWE emulation).

@@

where $\hat{\theta}_{RCT}$ and $\hat{\theta}_{RWE}$ are the treatment effect estimates (hazard ratios or median survival time differences) and $\text{Var}(\hat{\theta}_{RCT})$ and $\text{Var}(\hat{\theta}_{RWE})$ are the corresponding variances for RCT and RWE, respectively. The resulting SMDs will be interpreted such that with an SMD of 1.00, the effect estimate from the RCT and the RWE emulation are 1 standard deviation apart. For an $\alpha$-level of 0.05, the null hypothesis of no difference would be rejected whenever $|Z| > 1.96$.

@@

To establish an analytic cohort, key eligibility criteria will be applied in which patients with missing values in eligibility criteria are considered eligible in the respective attrition steps to allow for thorough missing data investigations. These missing data investigations will empirically assess assumptions on potential underlying missingness mechanisms according to Rubin’s classification of missing data (i.e., missing completely at random \[MCAR\], missing at random \[MAR\] and missing not at random \[MNAR\]).[@rubin1976inference] To that end, we will adopt a principled process on missing data that was developed as part of a FDA Sentinel Innovation Center causal inference workstream that empirically evaluates different aspects across partially observed covariates based on three group diagnostics.[@weberpals2024smdi; @weberpals2024] In brief, these diagnostics cover (1) comparisons of patients characteristics with and without an observed level of the partially observed covariate, (2) ability to predict missingness given observed data, and (3) assessments if outcomes between patients with a missing value are systematically different. Together with expert domain knowledge and assumptions about the underlying missing data structure through canonical causal diagrams,[@moreno2018canonical] this will inform decisions regarding the in- or exclusion of patients with missing values in key eligibility criteria and potential sensitivity analyses to assess the robustness of these decisions.[@tompsett2018use]

@@

Building on an established approach based on the RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized], the ENCORE project aims to emulate 12 oncology trials using multiple EHR data sources to inform the potential use of RWE for regulatory science in oncology. The project focuses on four common cancers and will evaluate the agreement of treatment effect estimates between RCTs and their respective emulations. Historically, administrative health claims databases have been the backbone of most research in RWE. With increasing access to EHR data and a maturing set of methodological approaches to draw causal inferences from such data, the ENCORE project is one of the first of its kind to evaluate when and how RWD can be used to deliver similar causal conclusions compared to RCTs in the field of oncology.

The RCT DUPLICATE project[@wang2023emulation; @franklin2020nonrandomized] identified multiple emulation challenges which we expect to also encounter in ENCORE. One particular aspect that we may not always be able to emulate is the exact distribution of patient characteristics of the trial population. This can be due to the lack of data granularity and comprehensiveness to emulate relevant eligibility criteria or the fact that large pivotal trials in oncology are typically conducted in multiple countries worldwide, which we will not be able to mirror given that all considered databases reflect the US only. This may be a critical factor especially given that the pathophysiology, prognosis and factors that drive heterogeneous treatment outcomes of certain cancers differ between countries. For example, some cancer indications (e.g., GI cancers) or some genetic mutations (e.g., EGFR mutations in lung adenocarcinoma) are much more prevalent in Asian countries compared to the US or Europe.

@@

This issue may be even augmented with more complex treatment protocols like combination regimens (e.g., CheckMate9LA) as compared to monotherapies (e.g., CheckMate057) since the ascertainment of the exposure needs to happen over a pre-defined time window which is often difficult to calibrate in RWD. As a result, exposure misclassification (due to overly ascertainment windows) and selection bias (due to overly long ascertainment windows) are common trade-offs in such scenarios. Alternative analytic approaches which target a per-protocol estimand, such as the clone-censor-weight design,[@gaber2024mystifying] may be viable options if the ITT estimand cannot be estimated due to the aforementioned parameters, provided that necessary covariate measurements are available to account for the artificial censoring introduced with these methods.

@@

```{r}
#| label: fig-initiators
#| fig-cap: "Example visualization of descriptive drug utilization analyses displaying a) initiation trends between compared regimens based on calendar time, b) cumulative rate of patients switching to another line of treatment."
#| fig-widh: 3
#| fig-height: 6
